Open Data, Open Source and Open Standards in chemistry - The Blue Obelisk five years on

### 2011-04-18 paper.tex
Added some brief information about Indigo, KNIME, and CurlySMILES

\author{Charles A Darwin\correspondingauthor$^{1,2}$%
       \email{Charles A Darwin\correspondingauthor - charles@londonzoo.co.uk}%
      \and 
         Egon L Willighagen$^3$%
         \email{Egon L Willighagen - egon.willighagen@ki.se}
     \and 
         Rajarshi Guha$^4$%
         \email{Rajarshi Guha - guhar@mail.nih.gov}
	 \and 
	     Christoph Steinbeck$^6$%
	         \email{Christoph Steinbeck - steinbeck@ebi.ac.uk}
     \and 
         Noel M O'Boyle$^5$%
         \email{Noel M O'Boyle - n.oboyle@ucc.ie}
     \and 
         Peter Murray-Rust$^7$%
         \email{Peter Murray-Rust - pm286@cam.ac.uk}
      \and
         Jane E Doe\correspondingauthor$^2$%
         \email{Jane E Doe\correspondingauthor - jane.e.doe@cambridge.co.uk}
      }

@@

\address{%
    \iid(1)Life Sciences Department, Kings College London, Cornwall House,%
        Waterloo Road, London, UK\\
    \iid(2)Department of Zoology, Cambridge, Waterloo Road, London, UK\\
    \iid(3)Division of Molecular Toxicology, Institute of Environmental Medicine, %
        Nobels vaeg 13, Karolinska Institutet, 171 77 Stockholm, Sweden\\
    \iid(5)Analytical and Biological Chemistry Research Facility, Cavanagh Pharmacy Building, University College Cork, College Road, Cork, Co. Cork, Ireland\\
    \iid(6)Cheminformatics and Metabolism Team, European Bioinformatics Institute (EBI), Wellcome Trust Genome Campus, Hinxton, Cambridge, UK
    \iid(7)Unilever Centre for Molecular Science Informatics, Department of Chemistry, University of Cambridge, Lensfield Road, CB2 1EW, UK
}%

@@

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
The Blue Obelisk movement was established in 2005 at the
229\textsuperscript{th} National Meeting of the American Chemistry
Society as a response to the lack of open data, open standards and
open source (ODOSOS) in chemistry. While other scientific disciplines
such as physics, biology and astronomy (to name a few) were embracing
new ways of doing science and reaping the benefits of community
efforts, there was little if any innovation in the field of chemistry
and scientific progress was actively hampered by the lack of access to
data and tools.
Since 2005 it has become evident that a good amount of development in open 
chemical information and information is driven by the demands of neighbouring
scientific fields. In many areas in biology, for example, the importance of 
small molecules and their interactions and reactions in biological systems
has been realised. In fact, one of the first free and open databases and ontologies 
of small molecules was created as a resource about chemical structure and nomenclature
by biologists \cite{DeMatos:2009p3839}.
Open workflow systems such as Taverna \cite{Hull:2006p60} have opened the door for 
an easy and seamless integration of free and open software (FOS) components into executable workflows 
or data pipelines. A combination of the Chemistry Development Kit and Taverna, for instance, was
reported in 2010 \cite{Kuhn:2010p4001}.

@@

Two new Open Source cheminformatics toolkits have appeared since the
original paper. In 2006 Rational Discovery, a cheminformatics service
company (since closed down), released RDKit \cite{WebRDKit} under the
BSD License. RDKit continues to be actively developed and includes
code donated by Novartis. More recently, GGA Software Services (a
contract programming company) released the Indigo toolkit
\cite{WebIndigo} and associated software in 2009 under the GPL. [More
text on Indigo here]

@@

@@

>>>>

\author{Charles A Darwin\correspondingauthor$^{1,2}$%
       \email{Charles A Darwin\correspondingauthor - charles@londonzoo.co.uk}%
      \and 
         Egon L Willighagen$^3$%
         \email{Egon L Willighagen - egon.willighagen@ki.se}
     \and 
         Rajarshi Guha$^4$%
         \email{Rajarshi Guha - guhar@mail.nih.gov}
	 \and 
	     Christoph Steinbeck$^6$%
	         \email{Christoph Steinbeck - steinbeck@ebi.ac.uk}
     \and 
         Noel M O'Boyle$^5$%
         \email{Noel M O'Boyle - n.oboyle@ucc.ie}
     \and 
         Peter Murray-Rust$^7$%
         \email{Peter Murray-Rust - pm286@cam.ac.uk}
      \and
         Jane E Doe\correspondingauthor$^2$%
         \email{Jane E Doe\correspondingauthor - jane.e.doe@cambridge.co.uk}
      \and
         Dmitry Pavlov$^8$%
         \email{Dmitry Pavlov - dpavlov@ggasoftware.com}
      }

@@

\address{%
    \iid(1)Life Sciences Department, Kings College London, Cornwall House,%
        Waterloo Road, London, UK\\
    \iid(2)Department of Zoology, Cambridge, Waterloo Road, London, UK\\
    \iid(3)Division of Molecular Toxicology, Institute of Environmental Medicine, %
        Nobels vaeg 13, Karolinska Institutet, 171 77 Stockholm, Sweden\\
    \iid(5)Analytical and Biological Chemistry Research Facility, Cavanagh Pharmacy Building, University College Cork, College Road, Cork, Co. Cork, Ireland\\
    \iid(6)Cheminformatics and Metabolism Team, European Bioinformatics Institute (EBI), Wellcome Trust Genome Campus, Hinxton, Cambridge, UK
    \iid(7)Unilever Centre for Molecular Science Informatics, Department of Chemistry, University of Cambridge, Lensfield Road, CB2 1EW, UK
    \iid(8)GGA Software Services LLC, 41 Nab. Chernoi rechki 194342, Saint Petersburg, Russia
}%

@@

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
The Blue Obelisk movement was established in 2005 at the
229\textsuperscript{th} National Meeting of the American Chemistry
Society as a response to the lack of open data, open standards and
open source (ODOSOS) in chemistry. While other scientific disciplines
such as physics, biology and astronomy (to name a few) were embracing
new ways of doing science and reaping the benefits of community
efforts, there was little if any innovation in the field of chemistry
and scientific progress was actively hampered by the lack of access to
data and tools.
Since 2005 it has become evident that a good amount of development in open 
chemical information and information is driven by the demands of neighbouring
scientific fields. In many areas in biology, for example, the importance of 
small molecules and their interactions and reactions in biological systems
has been realised. In fact, one of the first free and open databases and ontologies 
of small molecules was created as a resource about chemical structure and nomenclature
by biologists \cite{DeMatos:2009p3839}.
Open workflow systems such as Taverna \cite{Hull:2006p60} and KNIME\cite{WebKNIME}
have opened the door for 
an easy and seamless integration of FOS components into executable workflows 
or data pipelines. A combination of the Chemistry Development Kit and Taverna, for instance, was
reported in 2010 \cite{Kuhn:2010p4001}.

KNIME currently has a built-in collection of CDK-based and
OpenBabel-based nodes\cite{WebKNIMEExtensions} for implementing
chemistry workflows. Another collection of chemistry nodes---based on
the RDKit and its Java wrapper---appeared in 2010 and is available to
KNIME users from the KNIME's official ``Community Updates'' site.
Indigo nodes for KNIME are in preparation and will appear nearby the
RDKit nodes by May 2011.

@@

Two new Open Source cheminformatics toolkits have appeared since the
original paper. In 2006 Rational Discovery, a cheminformatics service
company (since closed down), released RDKit \cite{WebRDKit} under the
BSD License. RDKit continues to be actively developed and includes
code donated by Novartis.

More recently, GGA Software Services (a contract programming company)
released the Indigo toolkit \cite{WebIndigo} and associated software
in 2009 under the GPL. Indigo is written from scratch in C++ and has
high-level wrappers for using it in C, Java, Python, or .NET
environment. Like other toolkits, Indigo provides support for
tetrahedral ans cis-trans stereochemistry, 2D coordinate generation,
exact/substructure/SMARTS matching, fingerprint generation (all these
both for molecules and reactions), and canonical SMILES computation.
It also provides some less common functionality, like matching
tautomers ans resonance substructures, enumeration of subgraps,
finding maximum common substructure of $N$ input structures, and
enumerating reaction products. Indigo also has its rendering plugin,
which is used in a lot of commercial and open-sourse projects as the
main 2D chemical rendering engine.

@@

Also, Indigo is going to provide its own plugin with an interface to
the InChI implementation in May 2011. This will enable developers to
calculate InChI and InChIKey-s within Python and .NET projects (not to
mention Java), eliminating the trouble of compiling and using binary
modules directly.

@@

Recently proposed CurlySMILES\cite{CurlySMILES} is an extension of the
SMILES notation, which allows to define crystals structures, polymers,
electron delocalization charges, molecule interations, and many other
features absent in the initial SMILES specification. There has been
discussions about including parts of the CurlySMILES notation into
OpenSMILES, especially polymers.
